We tried to figure out why our project failed. In the process, we
characterised the differences between formal and implicit communication
channels. In every project---here, computer security testing---all the reports,
schedules, and emails comprise the formal channels through which
information exchanged amongst participants is supposed to flow. Formal
channels are invariably polite, wordy, one-to-many communications that tell
recipients nothing they didn't already know. In contrast, implicit channels
are profane, timely, idiomatic, and overwhelmingly tend to be one-to-one.
Implicit channels carry only important information. Necessary but not
sufficient, formal channels, by their nature, are highly resistant to
disruption. Implicit channels are fragile. Analogous to channel coding,
often deliberately unintelligible to an outsider, senders use phrasing,
metaphor, or sarcasm in a way that forces the recipient to work hard to decode the
meaning, thereby protecting the sender of information when risk from disclosure
is high. When placed against the inefficient but robust coding of formal
channels, shared context justifies both the relative fragility and
the practical necessity for implicit channels, thereby explaining their
emergence in many cases. Forerunner communication is the most valuable kind of
implicit channel; we found evidence of it being a diagnostic indicator of
project success. The fragility of implicit channels serves no direct
purpose, but the fact of their existence in spite of that drawback argues
for their necessity. Narrow-bandwidth and high-entropy, one reason for the
emergence of implicit channels now becomes clear: the insufficiency of
formal channels' capacity in light of typical schedule pressure (which
pressure directly leads to the emergence of implicit channels carrying
forerunner communication needed to meet the schedule). Like personally
identifiable information (PII), organisations do not get to choose the implicit
channels they get. Corporate email `data retention' policies (more like self-imposed,
voluntary lacunar amnesia) cause compensating implicit channels to come
into existence.

